Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-26 Manuscript_Nater_etal.qmd
Added missing references

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program. This often limits the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>). In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF). As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products. <!-- We might need to justify the latter a bit, as the rodent variance is calculated different than the other variances -->

@@

The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas. Consequently, many additional opportunities for improving and refining the modelling framework itself remain. First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys. In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances. When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e. that there is some observation error in the number reported. This could be incorporated by including an additional layer of hierarchy to the observation process (see e.g. REF), and possibly further extended to also account for error in judging the observation distance [e.g., @marques2004]. Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds. @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations. If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref). In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assingn "unknown" age class to juveniles than adults. One reason for this is that observers look for specific signs to classifying a bird as adult (e.g. size, male sound), and might default to juvenile or unknown if the signs are not clearly detected. Future studies should investigate to what degree available information on e.g. group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models (CARs). Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

>>>>

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program. This often limits the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence(-absence) or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates [@dickinson2010; @johnston2023]. The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>). In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF). As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

@@

Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$). To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$). The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products. This approach for variance decomposition is equivalent to that used by @nater2018 and inspired by @nakagawa2013.

@@

The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas. Consequently, many additional opportunities for improving and refining the modelling framework itself remain. First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys. In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances. When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e. that there is some observation error in the number reported. This could be incorporated by including an additional layer of hierarchy to the observation process [see e.g., @hamilton2018], and possibly further extended to also account for error in judging the observation distance [e.g., @marques2004]. Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds. @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations. If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations [@mccrea2013]. In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assingn "unknown" age class to juveniles than adults. One reason for this is that observers look for specific signs to classifying a bird as adult (e.g. size, male sound), and might default to juvenile or unknown if the signs are not clearly detected. Future studies should investigate to what degree available information on e.g. group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018]. Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
